---
title: "Addition to Austin Chan's Tibble Vignette"
author: "Stephen Jones"
date: "April 27, 2019"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

```{r}
library(tibble)
library(readr)

women_stem_tibble =  read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/women-stem.csv")

women_stem_tibble
```

```{r}
women_stem_df = read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/women-stem.csv")

women_stem_df
```


#Tidyverse, Part II, added by Stephen Jones.  

##Subsetting methods with tibbles  

Subsetting tibbles is accomplished using methods developed to subset dataframes; `$`, `[[`, or `[` may be used to call columns, rows, or cells.  

`which` may be implemented to subset on specific values:  

```{r message=FALSE,warning=FALSE}

#create a tibble of Engineering majors
WSeng_tibble <- women_stem_tibble[which(women_stem_tibble$Major_category == "Engineering"),] 

head(WSeng_tibble,5)

```  

Most commands are compatible with tibbles. Below, `grepl` is used to create a tibble of biological or biomedical majors.  


```{r message=FALSE,warning=FALSE}

field<-c("BIOLOGICAL","BIOMEDICAL")

#keep first five rows, columns 2 through 7
WSbio_tibble <- subset(women_stem_tibble, grepl(paste(field, collapse= "|"), women_stem_tibble$Major))


```

Rows and columns may be called by number:  

```{r message=FALSE,warning=FALSE}

#keep first five rows, columns 2 through 7
WSfirstfive_tibble <- women_stem_tibble[1:5,2:7] 

head(WSfirstfive_tibble,5)

```

Tibbles may be subset by detecting a substring with column names:  


```{r message=FALSE,warning=FALSE}

library(stringr)
#subset based on appearance of punctuation in column name using stringr
WSpunc_tibble<-women_stem_tibble[str_detect(names(women_stem_tibble), "[[:punct:]]")]

head(WSpunc_tibble,5)

```
Other options that may be used for subsetting:  

```{r message=TRUE,warning=TRUE}

# Subsetting single columns:
women_stem_tibble[, "Major_code"]
```

`drop=TRUE` coerces the result to a vector if a single column is called or coerces to a list if a single row is called; otherwise, the result is returned with the lowest possible dimension. The code chunk below returns a list.  


```{r message=TRUE,warning=TRUE}
women_stem_tibble[, "Major_code", drop = TRUE]

```

The commands in the following code chunk return lists:

```{r message=FALSE,warning=FALSE}
# Subset single rows with the drop:
women_stem_tibble[1, , drop = TRUE]

as.list(women_stem_tibble[1, ])

```

Tibbles are returned when `[` is used to specify columns. `[[` returns an atomic vector.  


```{r message=FALSE,warning=FALSE}
#Other options that return lists:
women_stem_tibble[1, ]

women_stem_tibble[1, c("Major", "Major_code")]

women_stem_tibble[, c("Major", "Major_code")]

women_stem_tibble[c("Major", "Major_code")]

women_stem_tibble["Major_code"]


#returns an atomic vector.
women_stem_tibble[["Major_code"]]


```





